### 2020-07-16 paper.tex
start implementation alg box
some tweaks to Section 3 and changes to Algortihm boxes

@@

Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two general measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$. 
Involutive MCMC implements a transition kernel that is invariant for a \textit{model distribution} given by $p : X \to [0, \infty)$, a probability density over $X$ with respect to $\mu_P$. Each iteration, the algorithm first samples auxiliary variables $y \in Y$ from an \emph{auxiliary distribution} $q_x$ based on the model's current state $x$: for each $x \in X$ such that $p(x) > 0$, $q_x : Y \to [0, \infty)$ is a probability density with respect to $\mu_Q$.
% We call $p$ the \emph{model density} and we call $q_x$ \emph{auxiliary densities}.

@@

Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$. Involutive MCMC runs $f$ on $(x, y)$ to obtain $(x', y')$, then computes an acceptance probability:
\begin{equation}
\alpha(x, y) = \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)
\end{equation}
With probability $\alpha$, the new state $x'$ is returned; otherwise, the previous state $x$ is repeated.
%Let $y \sim q_x(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_x(y) \mu_Q(dy)$ for $A \in \Sigma_Q$.

@@

The general measure-theoretic formulation of involutive MCMC above does not specify how to compute a Radon-Nikodym derivative, or the densities.
We now derive an algorithm for involutive MCMC when the model and auxiliary distributions are defined on spaces of dictionaries.

% TODO should we just segment the addresses into discrete and continuous addresses as part of the dictionary formalism above -- that would allow us to move this up. the added flexibility of full per-address measure is not used anyways..

Suppose that the model distribution $p$ and auxiliary distributions $q_x$ are all probability distributions on dictionaries.
Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}$, so that $Z = X \times Y \subseteq \mathcal{D} \times \mathcal{D}$.
%And suppose that there exists an involution $f : Z \to Z$ such that (i) 
Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $x_a$ and $y_a$ for all discrete keys $a$ are both constant (so that each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries).
Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
Let $e(z) \in E$ denote the element of the partition for an alement $z \in Z$.
Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
\begin{equation}
f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
\end{equation}
Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
Algorithm~\ref{alg:involutive-mcmc} defines involutive MCMC in this more structured setting, in which the Radon-Nikodym is replaced with a Jacobian (see Section~\ref{sec:radon-nikodym-special-case} for a derivation).

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{involutive-mcmc-kernel}{$p$, $q$, $f$, $x$}
    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary dictionary}
    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
    \State $\alpha \gets
        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)$
    \State $r \sim \mathrm{Uniform}(0, 1)$
    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure
\end{algorithmic}
\caption{Involutive MCMC for Dictionaries}
\label{alg:involutive-mcmc}
\end{algorithm}

@@

>>>>

% algorithmic
\algrenewcomment[1]{\(\triangleright\) #1}
\algnewcommand{\LineComment}[1]{\State \(\triangleright\) #1}
%\algblock[Foo]{execute}{end}

\algblockx[exec]{execute}{END}%
    [1][Unknown]{Execute #1}%
    [1][Unknown]{\vspace{-4mm}}

@@

Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two general measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$. 
Involutive MCMC (Algorithm~\ref{alg:involutive-mcmc}) implements a transition kernel that is invariant for a \textit{model distribution} given by $p : X \to [0, \infty)$, a probability density over $X$ with respect to $\mu_P$. Each iteration, the algorithm first samples auxiliary variables $y \in Y$ from an \emph{auxiliary distribution} $q_x$ based on the model's current state $x$: for each $x \in X$ such that $p(x) > 0$, $q_x : Y \to [0, \infty)$ is a probability density with respect to $\mu_Q$.
% We call $p$ the \emph{model density} and we call $q_x$ \emph{auxiliary densities}.

@@

Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$. Involutive MCMC runs $f$ on $(x, y)$ to obtain $(x', y')$, then computes an acceptance probability $\alpha$.
With probability $\alpha$, the new state $x'$ is returned; otherwise, the previous state $x$ is repeated.
%Let $y \sim q_x(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_x(y) \mu_Q(dy)$ for $A \in \Sigma_Q$.

@@

The general measure-theoretic formulation of involutive MCMC in Algorithm~\ref{alg:involutive-mcmc} does not specify how to compute the Radon-Nikodym derivative that is required for the acceptance probability.
When the algorithm is specialized for state spaces consisting of dictionaries, the acceptance probability simplifies and can be computed from standard Jacobian and probability density computations.
Because dictionaries are flexible objects, the algorithm remains quite general.

Suppose that the model distribution $p$ and auxiliary distributions $q_x$ are all probability distributions on dictionaries.
Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}$, so that $Z = X \times Y \subseteq \mathcal{D} \times \mathcal{D}$.
%And suppose that there exists an involution $f : Z \to Z$ such that (i) 
Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $x_a$ and $y_a$ for all discrete keys $a$ are both constant (so that each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries).
Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
Let $e(z) \in E$ denote the element of the partition for an alement $z \in Z$.
Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
\begin{equation}
f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
\end{equation}
Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
Then, the acceptance probability simplifies to:
\begin{equation}
\alpha(x, y) = \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
\end{equation}
Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{involutive-mcmc}{$p$, $q$, $f$, $x$}
    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary state}
    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
    \State $\alpha \gets
        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
    \State $r \sim \mathrm{Uniform}(0, 1)$
    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure
\end{algorithmic}
\caption{Involutive MCMC}
\label{alg:involutive-mcmc}
\end{algorithm}

@@

\begin{algorithm*}[h]
\begin{algorithmic}
\Procedure{auto-involutive-mcmc}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$, $x$}
    \State $(y, \log w_2) \gets \textproc{trace-and-score}(\mathcal{Q}_x)$ \Comment{Sample $y \sim q_x(\cdot)$ and compute log-density $\log q_x(y)$}
    \State $(x', y', \log D) \gets \textproc{run-involution}(\mathcal{F}, (x, y))$ \Comment{Compute $(x', y') = f(x, y)$, log Radon-Nikodym deriv. $\log D$}
    \State $\log w_1 \gets \textproc{score}(\mathcal{P}, x)$ \Comment{Compute log-density $\log p(x)$}
    \State $\log w_3 \gets \textproc{score}(\mathcal{P}, x')$ \Comment{Compute log-density $\log p(x')$}
    \State $\log w_4 \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ \Comment{Compute log-density $\log q_{x'}(y')$}
    \State $\alpha \gets \min\{1, \exp(\log w_3 - \log w_1 + \log w_4 - \log w_2 + \log J)\}$ \Comment{Compute acceptance probability}
    \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure\\

\begin{minipage}[t]{0.5\linewidth}
\Procedure{trace-and-score}{$\mathcal{P}$}
   \State $\log w \gets 0$
   \State $x \gets \{\}$
   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
        \State \hspace{5mm} 1. set $v \sim d$
        \State \hspace{5mm} 2. set $x[a] \gets v$
        \State \hspace{5mm} 3. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
        \State \hspace{5mm} 4. evaluate to $v$)
    \END
   \State \Return $(x, \log w)$
\EndProcedure
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\Procedure{score}{$\mathcal{P}$, $x$}
   \State $\log w \gets 0$
   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
        \State \hspace{5mm} 1. set $v \gets x[a]$
        \State \hspace{5mm} 2. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
        \State \hspace{5mm} 3. evaluate to $v$)
    \END
   \State \Return $\log w$
\EndProcedure
\end{minipage}\\

\Procedure{run-involution}{$\mathcal{F}$, $(x, y)$}
% TODO would be much simpler to describe if we didn't have to separate out x and y..
    \LineComment{Run the involution, and keep track of continuous addresses}
    \State $x' \gets \{\}$ \Comment{Initialize empty output dictionary}
    \execute[$\mathcal{F}$], but with
        \State $\mathtt{"write\_cont}(a, v)\mathtt{"} \equiv$ (1. set $x'[a] \gets v$ and 2. set $W \gets W \cup \{a\}$)
        \State $\mathtt{"read\_cont}(a)\mathtt{"} \equiv$ (1. set $R \gets R \cup \{a\}$ and 2. evaluate to $x[a]$)
        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (1. set $x'[a_2] \gets x[a_1]$ and 2. set $C \gets C \cup \{a_2\}$)
    \END
    \LineComment{Use automatic differentiation to compute Jacobian, skipping copied addresses}
    \For{$i$ in $1, \ldots, |W \setminus C|$}
        \State $J[i,:] \gets \nabla_{x_{R \setminus C}} \left( h_{e(x, y)}[(W \setminus C)_i] \right)$ \Comment{Gradient of $x'[(W \setminus C)_i]$ w.r.t. non-copied continuous inputs}
    \EndFor
    \State \Return $(x', \log |\det(J)|)$
\EndProcedure

\end{algorithmic}
\caption{Automated Involutive MCMC}
\label{alg:auto-involutive-mcmc}
\end{algorithm*}
